Discrete Quantum Theories and Computing

### 2018-03-11 dissertation.tex
dissertation.lyx "2.2.1 The geometry of entanglement" Updated...

\newcommand{\muB}{\ensuremath{\mu^{B}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{{#1}}}
\newcommand{\Fq}{\Fpx{q}}
\newcommand{\ff}[1]{\Fpx{#1}}
\newcommand{\ffzx}[2]{\Fpx{#1}^{{#2}\;*}}
\newcommand{\ffx}[2]{\Fpx{#1}^{{#2}}}
\newcommand{\ffzd}[1]{\ffzx{#1}{d}}
\newcommand{\ffd}[1]{\ffx{#1}{d}}
\newcommand{\VVec}[1]{\vec{\bf #1}}
\newcommand{\mathReal}{\mathbb{R}}
\newcommand{\mathComplex}{\mathbb{C}}
\newcommand{\braket}[2]{\ip{#1}{#2}}
\newcommand{\todo}[1]{\textbf{TODO.~#1}}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Sphere}[1]{\mathbf{S}^{#1}}
\newcommand{\R}{\mathReal}
\newcommand{\rme}{\mathrm{e}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\CP}[1]{\mathComplex\mathbf{P}^{#1}}  % Complex projective space
\newcommand{\DCP}[1]{\mathbf{D}\mathComplex\mathbf{P}^{#1}} % Discrete complex projective
\def\fh{\mathfrak{h}}
\newcommand{\uf}{U_{\!f}}
\newcommand{\scalarPlus}{+}
\newcommand{\boolt}{\textsf{Bool}} 
\newcommand{\bfalse}{\texttt{\textbf{false}}}
\newcommand{\btrue}{\texttt{\textbf{true}}}
\newcommand{\dotprod}{dot product}

@@

\begin{comment}
An $n$-qubit pure state in the conventional quantum theory (CQT)
is represented as a vector in the $2^{n}$-dimensional Hilbert space.
If we eliminate all symmetries, an irreducible state is actually a
point in the projective Hilbert space \cite{MosseriDandoloff2001,Jaeger2007}
also known as the complex projective space~$\CP{2^{n}-1}$ \cite{Hatcher2001,Bengtsson2007}.
These irreducible quantum states can also be classified as product
states and entangled states, and the latter one plays the essential
role for pure-state algorithms \cite{Mermin2007,Jaeger2007}.
\end{comment}
It has been 36 years since Feynman~\cite{Feynman1982Simulating}
proposed the idea about quantum computation. Although there are many
models for quantum computing, the circuit model is still one of the
most widespread models~\cite{544199} no matter whether in theoretical
study or in building a real quantum computation. Information in a
quantum circuit is stored in qubits. Each qubit is a two-dimensional
quantum system. For a two-qubit system, if the first and second qubit
are described as the state~$\ket{\psi}=\alpha_{0}\ket{0}+\alpha_{1}\ket{1}$
and $\ket{\phi}=\beta_{0}\ket{0}+\beta_{1}\ket{1}$, the state of
the whole system is their tensor product 
\begin{equation}
\begin{aligned} & \ket{\psi}\ket{\phi}=\ket{\psi}\otimes\ket{\phi}=\left(\alpha_{0}\ket{0}+\alpha_{1}\ket{1}\right)\otimes\left(\beta_{0}\ket{0}+\beta_{1}\ket{1}\right)\\
={} & \alpha_{0}\beta_{0}\ket{0}\otimes\ket{0}+\alpha_{0}\beta_{1}\ket{0}\otimes\ket{1}+\alpha_{1}\beta_{0}\ket{1}\otimes\ket{0}+\alpha_{1}\beta_{1}\ket{1}\otimes\ket{1}\\
={} & \alpha_{0}\beta_{0}\ket{0}\ket{0}+\alpha_{0}\beta_{1}\ket{0}\ket{1}+\alpha_{1}\beta_{0}\ket{1}\ket{0}+\alpha_{1}\beta_{1}\ket{1}\ket{1}
\end{aligned}
\end{equation}
in the four-dimensional Hilbert space. For a $n$-qubit system, if
we describe the $j$-th qubit as the state~$\ket{\psi_{j}}$, the
state of the whole system is tensor product of all subsystems $\ket{\Psi}=\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{j}}\otimes\cdots\otimes\ket{\psi_{n}}$
in the Hilbert space of dimension $2^{n}$.

@@

Entanglement may be regarded as one of the main characteristics distinguishing
quantum from classical mechanics. Entanglement involves quantum correlations
such that the measurement outcomes in one subsystem are related to
the measurement outcomes in another one. Within the standard framework,
given a quantum system composed of $n$ qubit subsystems, a pure state
of the total system $\ket{\Psi}$ is said to be entangled if it cannot
be written as a product of states of each subsystem. That is, a state
$\ket{\Psi}$ is entangled if $\ket{\Psi}\ne\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{j}}\otimes\cdots\otimes\ket{\psi_{n}}$,
where $\ket{\psi_{j}}$ refers to an arbitrary state of the $j$-th
qubit, and $\otimes$ represents the tensor product. This is equivalent
to saying that if one calculates the reduced density operator $\rho_{j}$
of the $j$th subsystem by tracing out all the other subsystems, $\rho_{j}={\sf tr}_{\{1,\cdots,j-1,j+1,\cdots,n\}}(\rho)$,
with $j=1,\cdots,n$ and $\rho=\ket{\Psi}\bra{\Psi}$, the normalized
state $\ket{\Psi}$, $\langle\Psi|\Psi\rangle=1$, is entangled if
and only if at least one subsystem state is \textit{mixed\/}; i.e.,
${\sf tr}_{j}(\rho_{j}^{2})<1$. For example, consider 
\begin{eqnarray}
\rho_{j}=\frac{1}{2}(\mathds{1}+\sum\limits _{\mu=x,y,z}\langle\sigma_{\mu}^{j}\rangle\ \sigma_{\mu}^{j}),
\end{eqnarray}
where $\sigma_{\mu}^{j}$, $\mu=x,y,z$, are the Pauli operators acting
on the $j$-th spin, 
\begin{equation}
\sigma_{\mu}^{j}=\overbrace{\mathds{1}\otimes\mathds{1}\otimes\cdots\otimes\underbrace{\sigma_{\mu}}_{j^{th}\ \mbox{factor}}\otimes\cdots\otimes\mathds{1}}^{n\ \mbox{factors}}\;,\label{pauli3}
\end{equation}
with 
\begin{eqnarray}
\hspace*{-0.5cm}\sigma_{x}=\left(\begin{array}{cc}
0 & 1\\
1 & 0
\end{array}\right),\sigma_{y}=\left(\begin{array}{cc}
0 & -i\\
i & 0
\end{array}\right),\sigma_{z}=\left(\begin{array}{cc}
1 & 0\\
0 & -1
\end{array}\right),
\end{eqnarray}
and $\langle\sigma_{\mu}^{j}\rangle=\langle\Psi|\sigma_{\mu}^{j}|\Psi\rangle$
denotes the corresponding expectation value. The vectors ${\bf X}_{j}=(\langle\sigma_{x}^{j}\rangle,\langle\sigma_{y}^{j}\rangle,\langle\sigma_{y}^{j}\rangle)$,
${\bf X}_{j}\in{\mathbb{R}}^{3}$, allow a geometric representation
of each reduced state in ${\mathbb{R}}^{3}$, satisfying $0\le\|{\bf X}_{j}\|\le1$.
Since ${\sf tr}_{j}(\rho_{j}^{2})=\frac{1}{2}(1+\|{\bf X}_{j}\|^{2})$,
the state $\ket{\Psi}$ is entangled if $\|{\bf X}_{j}\|<1$ for at
least one $j$, represented by a point {\em inside} the corresponding
local Bloch sphere. One may therefore consider $\ket{\Psi}$ to be
maximally entangled if $\|{\bf X}_{j}\|=0$ for all $j$. On the other
hand, the state $\ket{\Psi}$ is unentangled (i.e., a product state)
if $\|{\bf X}_{j}\|=1$ for all $j$, corresponding to points lying
on the surface of the Bloch sphere.

A natural geometric measure of multipartite entanglement is obtained
by defining the {\em purity of a state relative to a set of observables}
\cite{Barnum2003,Barnum2004}. If the set is chosen to be the set
of {\em all local observables}, i.e., corresponding to each of
the subsystems that compose the actual system, one recovers the standard
notion of entanglement for multipartite systems. For example, if the
system consists of $n$ qubits, we obtain a measure of conventional
entanglement by calculating the purity relative to the set $\fh=\{\sigma_{x}^{1},\sigma_{y}^{1},\sigma_{z}^{1},\ldots,\sigma_{x}^{n},\sigma_{y}^{n},\sigma_{z}^{n}\}$,
\begin{equation}
P_{\fh}=\frac{1}{n}\sum\limits _{j=1}^{n}\sum\limits _{\mu=x,y,z}\langle\sigma_{\mu}^{j}\rangle^{2}\:,\ \ 0\leq P_{\fh}\leq1\ .\label{purityMeasure.eq}
\end{equation}
Since $\fh$ is a semi-simple Lie algebra, its generalized unentangled
states are the generalized coherent states obtained by applying any
group operation to a reference state such as $\ket{\sf0}=\ket{0}\otimes\cdots\otimes\ket{0}$.
For the algebra $\fh$ of local observables, such group operations
are simply local rotations on each qubit. In other words, the group
orbit describing the generalized coherent states of $\fh$ comprises
all the product states of the form $\ket{\Psi}=\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{n}}$,
which have maximum purity (i.e., $P_{\fh}=1$). Other states such
as the Greenberger-Horne-Zeilinger state $\ket{\Psi}=\ket{{\sf GHZ}_{n}}=\frac{1}{\sqrt{2}}(\ket{0}\otimes\cdots\otimes\ket{0}+\ket{1}\otimes\cdots\otimes\ket{1})$
are (maximally) entangled relative to the set of local observables
(i.e., $P_{\fh}=0$).

Different entanglement measures are obtained when a set $\fh$ different
from the local observables is chosen. An obvious example, in particular,
is given by the set of all observables. In this case, the purity takes
its maximum value independently of the pure quantum state \cite{Barnum2003,Barnum2004},
expressing the fact that any state is a generalized coherent state
of the Lie algebra of all observables.

>>>>

\newcommand{\muB}{\ensuremath{\mu^{B}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{{#1}}}
\newcommand{\Fq}{\Fpx{q}}
\newcommand{\ff}[1]{\Fpx{#1}}
\newcommand{\ffzx}[2]{\Fpx{#1}^{{#2}\;*}}
\newcommand{\ffx}[2]{\Fpx{#1}^{{#2}}}
\newcommand{\ffzd}[1]{\ffzx{#1}{d}}
\newcommand{\ffd}[1]{\ffx{#1}{d}}
\newcommand{\VVec}[1]{\vec{\bf #1}}
\newcommand{\mathReal}{\mathbb{R}}
\newcommand{\mathComplex}{\mathbb{C}}
\newcommand{\braket}[2]{\ip{#1}{#2}}
\newcommand{\todo}[1]{\textbf{TODO.~#1}}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Sphere}[1]{\mathbf{S}^{#1}}
\newcommand{\R}{\mathReal}
\newcommand{\rme}{\mathrm{e}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\CP}[1]{\mathComplex\mathbf{P}^{#1}}  % Complex projective space
\newcommand{\DCP}[1]{\mathbf{D}\mathComplex\mathbf{P}^{#1}} % Discrete complex projective
\def\fh{\mathfrak{h}}
\newcommand{\uf}{U_{\!f}}
\newcommand{\scalarPlus}{+}
\newcommand{\boolt}{\textsf{Bool}} 
\newcommand{\bfalse}{\texttt{\textbf{false}}}
\newcommand{\btrue}{\texttt{\textbf{true}}}
\newcommand{\dotprod}{dot product}
\newcommand{\Tr}{\ensuremath{\mathop{\mathrm{Tr}}\nolimits}}

@@

\begin{comment}
An $n$-qubit pure state in the conventional quantum theory (CQT)
is represented as a vector in the $2^{n}$-dimensional Hilbert space.
If we eliminate all symmetries, an irreducible state is actually a
point in the projective Hilbert space \cite{MosseriDandoloff2001,Jaeger2007}
also known as the complex projective space~$\CP{2^{n}-1}$ \cite{Hatcher2001,Bengtsson2007}.
These irreducible quantum states can also be classified as product
states and entangled states, and the latter one plays the essential
role for pure-state algorithms \cite{Mermin2007,Jaeger2007}.
\end{comment}
It has been 36 years since Feynman~\cite{Feynman1982Simulating}
proposed the idea about quantum computation. Although there are many
models for quantum computing, the circuit model is still one of the
most widespread models~\cite{544199} no matter whether in theoretical
study or in building a real quantum computation. Information in a
quantum circuit is stored in qubits. Each qubit is a two-dimensional
quantum system. For a two-qubit system, if the first and second qubit
are described as the state~$\ket{\psi}=\alpha_{0}\ket{0}+\alpha_{1}\ket{1}$
and $\ket{\phi}=\beta_{0}\ket{0}+\beta_{1}\ket{1}$, the state of
the whole system is their tensor product 
\begin{equation}
\begin{aligned} & \ket{\psi}\ket{\phi}=\ket{\psi}\otimes\ket{\phi}=\left(\alpha_{0}\ket{0}+\alpha_{1}\ket{1}\right)\otimes\left(\beta_{0}\ket{0}+\beta_{1}\ket{1}\right)\\
={} & \alpha_{0}\beta_{0}\ket{0}\otimes\ket{0}+\alpha_{0}\beta_{1}\ket{0}\otimes\ket{1}+\alpha_{1}\beta_{0}\ket{1}\otimes\ket{0}+\alpha_{1}\beta_{1}\ket{1}\otimes\ket{1}\\
={} & \alpha_{0}\beta_{0}\ket{0}\ket{0}+\alpha_{0}\beta_{1}\ket{0}\ket{1}+\alpha_{1}\beta_{0}\ket{1}\ket{0}+\alpha_{1}\beta_{1}\ket{1}\ket{1}
\end{aligned}
\end{equation}
in the four-dimensional Hilbert space. For a $n$-qubit system, if
we describe the $j$-th qubit as the state~$\ket{\psi_{j}}$, the
state of the whole system is tensor product of all subsystems $\ket{\Psi}=\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{j}}\otimes\cdots\otimes\ket{\psi_{n}}$
in the Hilbert space of dimension $2^{n}$. However, not all quantum
system can be expressed as the tensor product of its subsystems. This
kind of system is called entangled, and will be described in the next
section.

@@

\todo{{Need to define }(partial) trace and mixed state before this
subsection...}

Entanglement may be regarded as one of the main characteristics distinguishing
quantum from classical mechanics. Entanglement involves quantum correlations
such that the measurement outcomes in one subsystem are related to
the measurement outcomes in another one. Within the standard framework,
given a quantum system composed of $n$ qubit subsystems, a pure state
of the total system $\ket{\Psi}$ is said to be entangled if it cannot
be written as a product of states of each subsystem. That is, a state
$\ket{\Psi}$ is entangled if $\ket{\Psi}\ne\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{j}}\otimes\cdots\otimes\ket{\psi_{n}}$,
where $\ket{\psi_{j}}$ refers to an arbitrary state of the $j$-th
qubit, and $\otimes$ represents the tensor product. This is equivalent
to saying that if one calculates the reduced density operator $\rho_{j}$
of the $j$-th subsystem by tracing out all the other subsystems,
$\rho_{j}=\Tr_{\{1,\cdots,j-1,j+1,\cdots,n\}}\left(\rho\right)$,
with $j=1,\cdots,n$ and $\rho=\op{\Psi}{\Psi}$, the normalized state~$\ket{\Psi}$
is entangled if and only if at least one subsystem state is \textit{mixed};
i.e., $\Tr_{j}\left(\rho_{j}^{2}\right)<1$. For example, consider
\begin{equation}
\rho_{j}=\frac{1}{2}\left(\sigma_{0}+\sum\limits _{\mu=x,y,z}\expval{\sigma_{\mu}^{j}}\sigma_{\mu}^{j}\right)\,,
\end{equation}
where $\sigma_{\mu}^{j}$, $\mu=x,y,z$, are the Pauli operators acting
on the $j$-th spin~\cite{544199}, 
\begin{equation}
\sigma_{\mu}^{j}=\overbrace{\sigma_{0}\otimes\cdots\otimes\sigma_{0}\otimes\underbrace{\sigma_{\mu}}_{j^{th}\ \mbox{factor}}\otimes\sigma_{0}\otimes\cdots\otimes\sigma_{0}}^{n\ \mbox{factors}}\,,\label{pauli3}
\end{equation}
with 
\begin{equation}
\begin{aligned}\sigma_{0} & =\left(\begin{array}{cc}
1 & 0\\
0 & 1
\end{array}\right)\,, & \sigma_{x} & =\left(\begin{array}{cc}
0 & 1\\
1 & 0
\end{array}\right)\,, & \sigma_{y} & =\left(\begin{array}{cc}
0 & -i\\
i & 0
\end{array}\right)\,, & \sigma_{z} & =\left(\begin{array}{cc}
1 & 0\\
0 & -1
\end{array}\right)\,,\end{aligned}
\end{equation}
and $\expval{\sigma_{\mu}^{j}}=\melem{\Psi}{\sigma_{\mu}^{j}}{\Psi}$
denotes the corresponding expectation value. The vectors 
\begin{equation}
{\bf X}_{j}=\left(\expval{\sigma_{x}^{j}},\expval{\sigma_{y}^{j}},\expval{\sigma_{y}^{j}}\right)\in\mathbb{R}^{3}
\end{equation}
allow a geometric representation of each reduced state in $\mathbb{R}^{3}$,
satisfying $0\le\left\Vert {\bf X}_{j}\right\Vert \le1$. Since $\Tr_{j}\left(\rho_{j}^{2}\right)=\frac{1}{2}\left(1+\left\Vert {\bf X}_{j}\right\Vert ^{2}\right)$,
the state $\ket{\Psi}$ is entangled if $\left\Vert {\bf X}_{j}\right\Vert <1$
for at least one $j$, represented by a point \emph{inside} the corresponding
local Bloch sphere. One may therefore consider $\ket{\Psi}$ to be
maximally entangled if $\left\Vert {\bf X}_{j}\right\Vert =0$ for
all $j$. On the other hand, the state $\ket{\Psi}$ is unentangled
(i.e., a product state) if $\left\Vert {\bf X}_{j}\right\Vert =1$
for all $j$, corresponding to points lying on the surface of the
Bloch sphere.

A natural geometric measure of multipartite entanglement is obtained
by defining the \emph{purity of a state relative to a set of observables}
\cite{BKOV2003,BKOSV2004}. If the set is chosen to be the set of
\emph{all local observables}, i.e., corresponding to each of the subsystems
that compose the actual system, one recovers the standard notion of
entanglement for multipartite systems. For example, if the system
consists of $n$ qubits, we obtain a measure of conventional entanglement
by calculating the purity relative to the set $\fh=\left\{ \sigma_{x}^{1},\sigma_{y}^{1},\sigma_{z}^{1},\ldots,\sigma_{x}^{n},\sigma_{y}^{n},\sigma_{z}^{n}\right\} $,
\begin{equation}
\begin{aligned}P_{\fh} & =\frac{1}{n}\sum\limits _{j=1}^{n}\sum\limits _{\mu=x,y,z}\expval{\sigma_{\mu}^{j}}^{2}\,, & 0\leq P_{\fh}\leq1\,.\end{aligned}
\label{purityMeasure.eq}
\end{equation}
Since $\fh$ is a semi-simple Lie algebra, its generalized unentangled
states are the generalized coherent states obtained by applying any
group operation to a reference state such as $\ket{\overline{0}}=\ket{0}\otimes\cdots\otimes\ket{0}$.
For the algebra $\fh$ of local observables, such group operations
are simply local rotations on each qubit. In other words, the group
orbit describing the generalized coherent states of $\fh$ comprises
all the product states of the form $\ket{\Psi}=\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{n}}$,
which have maximum purity (i.e., $P_{\fh}=1$). Other states such
as the Greenberger-Horne-Zeilinger state $\ket{\Psi}=\ket{\mathsf{GHZ}_{n}}=\frac{1}{\sqrt{2}}\left(\ket{0}\otimes\cdots\otimes\ket{0}+\ket{1}\otimes\cdots\otimes\ket{1}\right)$
are (maximally) entangled relative to the set of local observables
(i.e., $P_{\fh}=0$).

Different entanglement measures are obtained when a set $\fh$ different
from the local observables is chosen. An obvious example, in particular,
is given by the set of all observables. In this case, the purity takes
its maximum value independently of the pure quantum state \cite{BKOV2003,BKOSV2004},
expressing the fact that any state is a generalized coherent state
of the Lie algebra of all observables.
